Viscous Transport in Eroding Porous Media

### 2019-09-24 docs/transport/erosion_transport.tex
NM

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in many environmental and
industrial applications.  Depending on the application, length scales
can vary from $10^{-6}\mathrm{m}$ to
$10^{-1}\mathrm{m}$~\cite{mil-chr-imh-mcb-ped1998}, velocity scales can
be as small as $10^{-1}\mathrm{m/day}$~\cite{kut-scr-dav-ham1995}, \nick{and
the scales of a single porous media can range over several orders of
magnitude.}
%
\todo[inline]{Nick: I am confused by the last part of the sentence (highlighted in red). What `scale' are you referring to? Length? You already described the length scales in the first part of the sentence. Maybe you mean to distinguish between the grain length scales and the overall length scales of the system. If so, you can use adjectives like `grain' and `macroscopic' or `overall' to clarify that.} 
%
Numerical methods that \st{accurately address all of these} \nick{resolve this range of} scales offer the ability to: (i) characterize dispersion~\cite{saf1959}, (ii) quantify mixing~\cite{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}, and (iii) potentially develop meaningful constitutive relationships that link the microscopic and macroscopic realms~\cite{mil-chr-imh-mcb-ped1998}. 
%
\todo[inline]{Nick: I reworded the above sentence. See if you agree with my changes.}
%
Examples of coarse-grained models for porous media flow include
pressure-saturation-permeability
relationships~\cite{mil-chr-imh-mcb-ped1998}, permeability-porosity
relationships~\cite{dar-mcc1998, car1937, won-kop-tom1984}, first
passage time distributions~\cite{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosities~\cite{hak-com-den2019, mat-kha-koz2008,
dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}, and more.

Flow in porous media is further complicated when there is a two-way
coupling between the flow and the geometry. For example, the grains of a
porous media evolve in time in applications involving
melting~\cite{bec-vis1998, rycroft2016asymmetric, morrow2019moving}, dissolution~\cite{kan-zha-che-he2002, mac2015shape, wykes2018self},
deposition~\cite{joh-eli1995}, biofilm growth~\cite{tan-val-wer2015}, and \nick{crack formation} \cite{cho2019crack}.  
%
\todo[inline]{Nick: I added some references. Check out the last one on crack formation in shrinkable, granular packings. It is really cool work done by an excellent experimentalist Sujit Datta. I met him and he voiced a strong interest in our work, so wanted to make you aware.}
%
We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical and hydrological
applications.  When a porous geometry erodes, certain qualitative
characteristics are unveiled that affect transport through the geometry.
For example, an eroded geometry often contains channels of high porosity
that connect the inlet to the outlet, and this results in velocities
that vary over several orders of magnitude~\cite{all-hea-lab-rei2002}.
Moreover, channelized geometries are heterogeneous and anisotropic, and
this affects the transport of quantities such as
contaminants~\cite{cve-che-wen1996, dag1987, kon-bre1978} and
heat~\cite{nil-sto1990, ree-sto1995}.  To characterize transport in
eroded geometries, we complete two undertakings: forming realistic
eroded porous geometries and characterizing transport through these
geometries.  The modelling effort and numerical methods used to simulate
erosion are largely based on our previous work~\cite{qua-moo2018}.
However, we improve our numerical methods so that simulations of more
realistic dense suspensions of eroding grains are possible
(Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  To
characterize transport in eroded porous geometries, we apply statistical
methods to develop coarse-grained variables.

@@

Because of the length and velocity scales in hydrological applications
such as groundwater flow~\cite{bea1972}, we model the hydrodynamics with
the two-dimensional incompressible Stokes equations.  To couple erosion
with the fluid, the grains are eroded at a rate that is proportional to
the magnitude of the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, par-izu2000}.  Since the vorticity and the
shear stress are equivalent on the grains in the zero Reynolds number
limit, we also compute the vorticity in the fluid bulk
(Figure~\ref{fig:Eroding50vort}).  A variety of numerical methods for
the fluid solver are available including lattice-gas or lattice
Boltzmann~\cite{kop-kat-tim1996, dar-mcc1998}, finite
volumes~\cite{suo-liu-gan2019, den-ica-hid2018,
sie-ili-pri-riv-gua2019}, finite differences~\cite{leb-ded-dav-bou2007,
knu-car2005}, and cellular automata~\cite{rot1988}.  However, simulating
erosion requires a method that is high-order, can resolve
nearly-touching arbitrarily shaped grains, and can accurately compute
derivatives of the velocity field.  Since the governing equations are
linear and homogeneous, we address these challenges by converting the
fluid equations to a boundary integral equation (BIE).

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in many environmental and
industrial applications.  Depending on the application, length scales
can vary from $10^{-6}\mathrm{m}$ to
$10^{-1}\mathrm{m}$~\cite{mil-chr-imh-mcb-ped1998}, velocity scales can
be as small as $10^{-1}\mathrm{m/day}$~\cite{kut-scr-dav-ham1995}, \awk{and
the scales of a single porous media can range over several orders of
magnitude.}
%
\todo[inline]{Nick: I am confused by the last part of the sentence (highlighted in purple). What `scale' are you referring to? Length? You already described the length scales in the first part of the sentence. Maybe you mean to distinguish between the grain length scales and the overall length scales of the system. If so, you can use adjectives like `grain' and `macroscopic' or `overall' to clarify that.} 
%
Numerical methods that \st{accurately address all of these} \nick{resolve this range of} scales offer the ability to: (i) characterize dispersion~\cite{saf1959}, (ii) quantify mixing~\cite{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}, and (iii) potentially develop meaningful constitutive relationships that link the microscopic and macroscopic realms~\cite{mil-chr-imh-mcb-ped1998}. 
%
\todo[inline]{Nick: I reworded the above sentence. See if you agree with my changes.}
%
Examples of coarse-grained models for porous media flow include
pressure-saturation-permeability
relationships~\cite{mil-chr-imh-mcb-ped1998}, permeability-porosity
relationships~\cite{dar-mcc1998, car1937, won-kop-tom1984}, first
passage time distributions~\cite{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosities~\cite{hak-com-den2019, mat-kha-koz2008,
dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}, and more.

Flow in porous media is further complicated when there is a two-way
coupling between the flow and the geometry. For example, the grains of a
porous media evolve in time in applications involving melting~\cite{bec-vis1998, rycroft2016asymmetric, favier2019rayleigh, morrow2019moving}, dissolution~\cite{kan-zha-che-he2002, mac2015shape, moore2017riemann, wykes2018self}, deposition~\cite{joh-eli1995, hewett2018modelling}, biofilm growth~\cite{tan-val-wer2015}, and \nick{crack formation} \cite{cho2019crack}.  
%
\todo[inline]{Nick: I added some references. Check out the last one on crack formation in shrinkable, granular packings. It is really cool work done by an excellent experimentalist Sujit Datta. I met him and he voiced a strong interest in our work, so wanted to make you aware.}
%
We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical, hydrological, \nick{and industrial} 
applications \cite{berhanu2012shape, hewett2017evolution, lopez2018cfd, allen2019sde, amin2019role}.
\todo[inline]{Nick: Check out the first reference, Berhanu, Kudrolli, Rothman PRE 2012. Maybe we can cite it a couple more times throughout since it has both the words `erosion' and `granular bed' in the title :)}
When a porous geometry erodes, certain qualitative characteristics are unveiled that affect transport through the geometry.
For example, an eroded geometry may contain channels of high porosity, \nick{which, though few in number and modest in volume fraction, transmit a large portion of the flux}. This arrangement results in velocities
that vary over several orders of magnitude~\cite{all-hea-lab-rei2002}.
%
\todo[inline]{Nick: I rearranged the above two sentences. See if you like it.}
%
Moreover, \nick{channelization creates a heterogeneous and anisotropic medium properties}, \st{and
this affects} \nick{affecting} the transport of \st{quantities} \nick{tracers} such as
contaminants~\cite{cve-che-wen1996, dag1987, kon-bre1978} and
heat~\cite{nil-sto1990, ree-sto1995}.  \awk{To characterize transport in
eroded geometries, we complete two undertakings: forming realistic
eroded porous geometries and characterizing transport through these
geometries}.  
\todo[inline]{Nick: The purple sentence is redundant: `To \underline{characterize transport} ... we undertake ... \underline{characterizing transport} through these geometries. Maybe try something like this:}
\nick{This study consists of two main undertakings: First, high-fidelity simulations of porous-medium erosion, and, second, the characterization of transport through the resulting geometries.}
Our modeling efforts build on previous work \cite{ris-moo-chi-she-zha2012, moore2013self, moore2017riemann}, in particular recent numerical methods developed to simulate erosion in the regime of Stokes flow \cite{qua-moo2018}. We, however, make a few key improvements to the numerical methods from \cite{qua-moo2018} to enable simulations of more realistic, dense suspensions of eroding grains (Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  To characterize transport in the resulting geometries, \st{we apply statistical methods to develop coarse-grained variables} \nick{examine coarse-grained variables through statistical analysis of tracer trajectories}.
\todo[inline]{I'm not sure about my modification above, but see if it gets across the message that was intended.}

@@

\st{Because of the length and velocity scales in hydrological applications
such as groundwater flow} \nick{Owing to the scales present in groundwater flow} \cite{bea1972}, we model the hydrodynamics with
the two-dimensional incompressible Stokes equations.  \st{To couple erosion
with the fluid} Meanwhile, individual grains erode at a rate \st{that is} proportional to \st{the magnitude of} the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, moore2013self, par-izu2000}.  Since the vorticity and the shear stress are equivalent on the grains in the zero Reynolds number
limit, we also compute the vorticity in the fluid bulk \nick{for visualization purposes} (Figure~\ref{fig:Eroding50vort}).  A variety of numerical methods for
the fluid solver are available including lattice-gas or lattice
Boltzmann~\cite{kop-kat-tim1996, dar-mcc1998}, finite
volumes~\cite{suo-liu-gan2019, den-ica-hid2018,
sie-ili-pri-riv-gua2019}, finite differences~\cite{leb-ded-dav-bou2007,
knu-car2005}, and cellular automata~\cite{rot1988}.  However, simulating
erosion requires a method that is high-order, can resolve
nearly-touching arbitrarily shaped grains, and can accurately compute
derivatives of the velocity field.  Since the governing equations are
linear and homogeneous, we address these challenges by converting the
fluid equations to a boundary integral equation (BIE).
